The state of knowledge for marine invertebrate biodiversity in the continental US

### 2018-05-01 manuscript.Rmd
include GP  comments

Given their diversity, marine invertebrates are underrepresented in iDigBio. While marine invertebrates are estimated to represent between 11% and 18% of the described global diversity (number of described marine inverterbrates species: 200,000-210,000 [@Appeltans2012], total number of described species: 1.2-1.9 million [@Mora2011,@Chapman2009]),
marine invertebrates from the US EEZ only make up `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$total_no_fossil*100)`% of the total number of non-fossil iDigBio records (`r round(idigbio_stats$total_no_fossil/1e6, 1)` million), and `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US records (`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million).

@@

This underrepresentation is also manifest in the intensity of the collecting effort. Marine invertebrates are estimated to represent  `r total_invert_prop`% of the total marine diversity[@worms-stats]. In OBIS records for the US EEZ, marine invertebrates represent `r invert_props$prop_obis_invert_samples`% of the occurrences but `r invert_props$prop_obis_invert_species`% of the total number of species; and in iDigBio, they represent `r invert_props$prop_idigbio_invert_samples`% of the records but `r invert_props$prop_idigbio_invert_species`% of the species. In contrast, marine vertebrates represent `r invert_props$prop_idigbio_vert_samples`% of the samples and `r invert_props$prop_idigbio_vert_species`% of the species in iDigBio, and `r invert_props$prop_obis_vert_samples`% of the samples and `r invert_props$prop_obis_vert_species`% of the species in OBIS (Fig. \@ref(fig:kingdom-diversity-plot)). In other words, while marine invertebrates are represented at a level on par with their diversity within the US EEZ in the databases, each species is represented by fewer samples than for vertebrates (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs._ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).

To assess the level of digitization of the fauna, we compared the species included in iDigBio and OBIS with the comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] that are available for the United States. For molluscs, of the `r format(itis_stats$moll_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS. For crustaceans, of the `r format(itis_stats$crust_nspp, big.mark=",")` species included in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS. Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (Fig.  \@ref(fig:idigbio-through-time)), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally. Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates in US waters. The higher overlap between the comprehensive checklist for  crustaceans and OBIS is attributable to occurences of planktonic species that are more represented in OBIS data than in museum collections.

The faunistic list we assembled represent a subset of the species that have been digitized, because only records that are associated with geographic coordinates are included. All records published by OBIS have geographic coordinates associated with them. To estimate how records with missing geographic coordinates in iDigBio affect the completeness of our faunistic list, we queried iDigBio using the comprehensive lists for molluscs and crustaceans, and calculated the proportion of species with no iDigBio records within the US EEZ, and which had "United States" or "USA" in the country field, but had no geographic coordinates associated with them. These species may occur within the limits of the US EEZ but would not be returned in the queries used to assemble the faunistic list. We found that `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo * 100)`% ($n=$ `r format(itis_stats$missing_geo_moll$n_spp, big.mark=",")`) of the species included in the comprehensive checklist for molluscs have records in iDigBio that place them in the United States based on the "Country" field, but do not have geographic coordinates associated with them that would ensures they fall within the limits of the EEZ. This percentage decreases to `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo_no_obis * 100)`% once we exclude species that have informed geographic coordinates in OBIS. For crustaceans, these proportions are lower at `r format_output(itis_stats$missing_geo_crust$prop_us_no_geo * 100)`% and `r format_output(itis_stats$missing_geo_crust$prop_us_no_geo_no_obis * 100)`% ($n=$ `r format(itis_stats$missing_geo_crust$n_spp)`) respectively. This means ...

@@

More records have been collected from the Atlantic Ocean, than the Gulf of Mexico, and the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively). The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species). Most samples have been collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between 0-200m. The higher species richness measured on the East coast is in part attributable to the wider continental shelf, the presence of more institutions, and a higher level of digitization of collections on the East coast (Table \@ref(tab:records-table)).

Spatial distribution of species richness with latitude differ between the Pacific and the Atlantic coasts. On the Pacific coast, the recorded richness is lowest around 40° of latitude, whereas on the Atlantic coast, it is highest at the same latitude (Fig. \@ref(fig:map-diversity-sampling)B, Fig. \@ref(fig:latitudinal-gradient)B). These trends are however strongly driven by sampling effort (Fig. \@ref(fig:latitudinal-gradient)C). Notably, for the Pacific coast, there is a deficit in the number of samples between 40° and 45° of latitude corresponding to the state of Oregon. In contrast, the peak in number of species observed and in the number of records around 42° on the Atlantic coast can be attributed to the presence of the Woods Hole Oceanographic Institute.

```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA (represented in light gray). (\\#fig:map-diversity-sampling)'}
knitr::include_graphics("figures/combined_map_sampling.pdf")
```

@@

```{r correlation-sampling, fig.cap='Relationship between number of records and number of species for each cell of the map used in Fig. \\@ref(fig:map-diversity-sampling). The black line represents the 1:1 relationship. (\\#fig:correlation-sampling)'}
knitr::include_graphics("figures/plot_sampling_effort.pdf")
```

@@

Spatial patterns of species richness are however strongly confounded by sampling effort. There is a tight correlation between the number of samples and number of species per cell used in Fig. \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r format(stat_sampling_effort$n, big.mark=",")`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig. \@ref(fig:correlation-sampling)). For most of the US EEZ, the sampling effort is insufficient to capture acurately the species diversity.

@@

Another indication of the inadequacy of the sampling is the relationship between the number of collected specimens and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Explorations in the late nineteenth century led to an exponential growth in the knowledge of the fauna, but collecting then tapered off, until the 1970s when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowed pace of collection activitity (and therefore a reduction in the number of specimens added) has been accompanied by a slow-down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` were documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^).

@@

Biodiversity surveys conducted under the auspices of the Bureau of Land Management (BLM) and the Minerals Management Service (MMS) have contributed significantly to our knowledge of the geographic distribution of marine invertebrates in US waters. In our dataset, we identified at least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r format_output(sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100)`% of all US marine invertebrates in iDigBio) from these surveys, `r format_output(prop_mms_75_85 * 100)`% of which were collected between 1975 and 1985. Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name)) %>% format(big.mark=",")` species collected as part of these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.

The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), suggest that any new sampling effort will lead to new observations and the discovery of new species.

@@

Further biases exist across phyla and body size. Large-bodied and easily accessible (e.g. dried specimens like shells, corals, echinoderms) specimens are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna. Of `r n_idigbio_phyla` phyla represented in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records. While these phyla represent
`r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species, they make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records. This proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` other phyla. Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012]. Additionally, the ecological implications of changes in the structure and composition of the communities of these small-bodied organisms, on higher trophic levels and ecosystems are largely unexplored.

The proportion of samples that are identified to species in museum collections show a marked decline in recent decades in most taxa other than mollusks, demonstrating limiting taxonomic effort and expertise (Fig. \@ref(fig:identification-through-time)). There is a marked break in the central trend around the 1970's indicating that it now takes decades, on average, to identify samples deposited in collections. For diverse groups such as annelids or arthropods, the proportion of specimens identified to the species-level is now below 30%.

@@

We extracted `r format(nrow(gom_worms_species), big.mark=",")` and `r format(nrow(kozloff_worms_species), big.mark=",")` marine invertebrate species from species checklists from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1996] respectively. For the Gulf of Mexico, biodiversity databases included more species for the same geographic area as the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW.

@@

Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases. The first category of biases is linked to the likelihood of having specimens for these species in digitized museum collections. Rare species may only have been collected a few times, and the corresponding records have not been digitized yet. Species that are difficult to preserve because they are large, very small, or fragile, are deposited in museums less frequently. The other category of biases is associated with data quality for these records. Some of these species may be in the databases but do not have geographic coordinates associated with them, and therefore would not be included in the results of our query. While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates are affecting the discrepancies between the checklists and the databases.

Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for those that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_eez*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_eez, big.mark=",")`) fall within US waters. Additional digitization of collections, including the capture of their geographic coordinates information, is needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems. Similarly, only a small fraction of the marine invertebrates records from institutions located on the Pacific coast has been digitized, as indicated by the paucity of Pacific coast collections listed in Table \@ref(tab:records-table).

Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional checklists: (1) the species have been documented for the first time since the publication of the checklist, and (2) the species records were not accessible or known to the authors.

Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico, as `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist),  have been collected for the first time since the publication of the checklist, but the results are more significant for the Pacific Northwest with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist).

The authors of these checklists did not have access to digital databases allowing them to access  museum records across institutions, which explains why some species were not included in the checklists. Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists. We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest.

@@

Since its introduction in the early 2000s, DNA barcoding has revitalized invertebrate taxonomy by facilitating species delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of specimens in environmental samples (eDNA) cost effectively [@Leray2015], making it possible to estimate diversity and identify taxa from species-rich samples at unprecedented scales. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies may soon be used routinely to monitor biodiversity continuously [@Bohan2017]. However, the utility of the data generated by these projects is limited by the paucity of reliably identified DNA barcode sequences that can serve for matching and identification of eDNA transcripts [@Leray2015]. Thus insights from eDNA technologies depend on the level of completeness and accuracy of comparative, species-level DNA barcode coverage.

The marine invertebrates of the US have generally been characterized genetically at a higher level than other parts of the world: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(combined_species), big.mark=",")`) of the digitized fauna has DNA barcodes available.  In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).  This proportion varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% for `r stat_barcoding$min_eez_phylum` to `r format_output(stat_barcoding$max_eez_value * 100)`% for `r stat_barcoding$max_eez_phylum` (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable coverage in level of DNA barcoding of the fauna to the EEZ, while the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by active Canadian barcoding efforts in British Columbia.

>>>>

Marine invertebrates are underrepresented in iDigBio compared to other taxa. Marine invertebrates represent 11% of described global species diversity (202,000 [@worms-stats] out of 1.7 million species [catalog-of-life]), but only `r format_output(nrow(idigbio_us_inverts_records_species)/idigbio_stats$eez_no_fossil*100)`% of the US non-fossil occurrence records ((`r round(idigbio_stats$eez_no_fossil/1e6, 1)` million) in iDigBio.

@@

Invertebrates are also represented in the marine data. About `r total_invert_prop`% of described marine species are invertebrates. The proportional species diversity of invertebrates is captured relatively effectively among US marine EEZ records in OBIS (`r invert_props$prop_obis_invert_species`%) and iDigBio (`r invert_props$prop_idigbio_invert_species`%). However, invertebrate occurrence records are disproportionately low: `r invert_props$prop_obis_invert_samples`% in OBIS and `r invert_props$prop_idigbio_invert_samples`% in iDigbio. Marine vertebrate species have three times as many occurrence records on average in these databases than invertebrate species (median number of records per species for invertebrates `r median_idigbio_invert` and `r median_obis_invert`, _vs._ and `r median_idigbio_vert` and `r median_obis_vert` for vertebrates, for iDigBio and OBIS respectively).

To assess how well documented species diversity is captured digitally, we compared the species retrieved from iDigBio and OBIS against comprehensive checklists for molluscs [@turgeon1998] and crustaceans [@mclaughlin2005] of the United States. Of the `r format(itis_stats$moll_nspp, big.mark=",")` mollusk species included in the checklist, `r itis_stats$prop_moll_idig`% are in iDigBio and `r itis_stats$prop_moll_obis`% are in OBIS. Of the `r format(itis_stats$crust_nspp, big.mark=",")` crustacean species in the checklist, `r itis_stats$prop_crust_idig`% are in iDigBio and `r itis_stats$prop_crust_obis`% are in OBIS. Given that molluscs and crustaceans are among the most sampled taxa, and that they have received more taxonomic attention than other phyla (Fig.  \@ref(fig:idigbio-through-time)), these percentages offer a conservative estimate for the fraction of the fauna that is available digitally. Therefore, using a conservative estimate of a 60% digitization level of the fauna, we can estimate that there is at least 29,000 species of marine invertebrates recorded in US waters. The higher overlap between the comprehensive checklist for crustaceans in OBIS than in iDigBio is attributable to planktonic species that are better represented in OBIS data than in museum collections.

The faunistic list we assembled represent a subset of the species that have been digitized, because only records that are associated with geographic coordinates are included. All records published by OBIS have geographic coordinates associated with them. To estimate the proportion of species missed because they lacked geographic coordinate in iDigBio, we queried iDigBio using the comprehensive list for molluscs and crustaceans, and calculated the proportion of species with no iDigBio records within the US EEZ, but which had "United States" or "USA" in the country field. These species should occur within within the limits of the US EEZ but would not be returned in the queries used to assemble the faunistic list. We found that `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo * 100)`% ($n=$ `r format(itis_stats$missing_geo_moll$n_spp, big.mark=",")`) of the species included in the comprehensive checklist for molluscs have records in iDigBio that place them in the United States based on the "Country" field, but do not have geographic coordinates associated with them that would ensures they fall within the limits of the EEZ. This percentage decreases to `r format_output(itis_stats$missing_geo_moll$prop_us_no_geo_no_obis * 100)`% once we exclude species that have informed geographic coordinates in OBIS. For crustaceans, these proportions are lower at `r format_output(itis_stats$missing_geo_crust$prop_us_no_geo * 100)`% and `r format_output(itis_stats$missing_geo_crust$prop_us_no_geo_no_obis * 100)`% ($n=$ `r format(itis_stats$missing_geo_crust$n_spp)`) respectively. This means ...

@@

The Atlantic Ocean and the Gulf of Mexico are much better doocumented in these databases than the Pacific Ocean (`r format(sum(combined_records$is_east_coast), big.mark=",")`, `r format(sum(combined_records$within_gom), big.mark=",")`, and `r format(sum(combined_records$is_west_coast), big.mark=",")` records respectively). The highest species richness has been recorded from the Gulf of Mexico (`r format(n_spp_gom, big.mark=",")` species), followed by the Atlantic Ocean (`r format(n_spp_east, big.mark=",")` species), and the Pacific Ocean (`r format(n_spp_west, big.mark=",")` species). The higher species richness documented on the East coast partly reflects sampling bias, reflected in many more records, partly attributable to the presence of more and older research institutions, higher levels of digitization (Table \@ref(tab:records-table)), and wider continental shelf there. Most samples have been collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A), with `r prop_idigbio_continental_shelf`% (iDigBio) and `r prop_obis_continental_shelf`% (OBIS) of the records collected between $leq$200m.

Sampling bias is further reflected in the lack of clear latitude diversity gradient (well documented regionally e.g. [@macpherson2002,@roy2000]) and differences between the Pacific and the Atlantic coasts. On the Pacific coast, the recorded richness is lowest around 40° of latitude, whereas on the Atlantic coast, it is highest at the same latitude (Fig. \@ref(fig:map-diversity-sampling)B, Fig. \@ref(fig:latitudinal-gradient)B). These trends are strongly influenced by sampling effort (Fig. \@ref(fig:latitudinal-gradient)C). Notably, for the Pacific coast, there is a deficit in the number of samples and species between 40° and 45° of latitude corresponding to the Oregon coast. In contrast, a peak in species and records around 42° on the Atlantic coast may be attributed to the Woods Hole Oceanographic Institute.

```{r map-diversity-sampling, fig.cap='Spatial distribution of the number of samples (A) and species richness (B) for marine invertebrates based on records retrieved from iDigBio and OBIS within the limits of the Exclusive Economic Zone for the continental USA (represented in light gray) in 0.2 x 0.2 degree cells. (\\#fig:map-diversity-sampling)'}
knitr::include_graphics("figures/combined_map_sampling.pdf")
```

@@

```{r correlation-sampling, fig.cap='Relationship between number of records and number of species per 0.2 x 0.2 degree cell in US EEZ (Fig. \\@ref(fig:map-diversity-sampling)). The black line represents the 1:1 relationship. (\\#fig:correlation-sampling)'}
knitr::include_graphics("figures/plot_sampling_effort.pdf")
```

@@

Further indication of sampling bias is a tight correlation between the number of samples and number of species per cell used in Fig. \@ref(fig:map-diversity-sampling), and `r format_output(stat_sampling_effort$p * 100)`% (n = `r format(stat_sampling_effort$n, big.mark=",")`) of the cells with $\leq$ 500 records have a number of species within 10% of the number of records (Fig. \@ref(fig:correlation-sampling)). For most of the US EEZ, the sampling effort is insufficient to capture the species diversity acurately.

@@

@@

@@

Another indication of the vagaries and inadequacy of the sampling is shown in the temporal record of collected species and documented species (Fig. \@ref(fig:idigbio-through-time)). Two time periods stand out: the 1870-80's and the 1960-80's. The first represents intensive efforts to document the US marine biota during the formative stages of the major museums and research organizations focused on marinie life, while the second is in large part the result of the biodiversity surveys conducted under the auspices of the Bureau of Ocean Energy Management (BOEM; formerly part of Bureau of Land Management, BLM, and the Minerals Management Service, MMS). At least `r format(sum(idigbio_mms_summary_by_year$n), big.mark=",")` records (or `r format_output(sum(idigbio_mms_summary_by_year$n)/nrow(idigbio_us_inverts_records_all)*100)`% of all US marine invertebrates) in iDigBio were from these surveys, `r format_output(prop_mms_75_85 * 100)`% of them collected between 1975 and 1985. Of the `r idigbio_mms_records %>% filter(rank %in% c("Species", "Subspecies")) %>% summarize(n = n_distinct(worms_valid_name)) %>% format(big.mark=",")` species collected in these surveys, `r nrow(idigbio_mms_species_only)` have not been collected outside of these surveys.

Since the BLM-MMS surveys, the rate of collection and documentation of new records have both slowed. During the 1960-1970 ~XX records were added per year, averaging `r round(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^ and 
`r round(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^ since 2000.

<!--- Note also difference in evenness across phyla – mollusks much more even, annelids most jerky – reflects broad vs focused collecting efforts].  Note however that the number of new records per sample taken has not declined appreciably (X to Y) in this period -->

The strong correlations between (1) the number of samples collected and the species richness (Fig. \@ref(fig:correlation-sampling)), and (2) the number of specimens collected per year and the number of species recorded for the first time (Fig. \@ref(fig:idigbio-through-time)C), also suggest that much of the US marine biodiversity remains undocumented.

Further biases exist across phyla and body size. Large-bodied and easily accessible (e.g. dried specimens like shells, corals, echinoderms) specimens are over-represented in museum collections, compared with small less accessible organisms, such as the meiofauna. Of `r n_idigbio_phyla` animal phyla represented from US waters in iDigBio, `r records_rare_phyla$n_rare_phyla` have <500 records. While these phyla represent
`r records_rare_phyla$prop_spp_rare_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_rare_phyla_collected, big.mark=",")` total species) of the global number of species <!-- source unclear -->, they make up `r records_rare_phyla$prop_recs_rare_phyla`% of the iDigBio records. This proportion rises to `r records_rare_phyla$prop_spp_common_phyla_collected`% (n = `r format(records_rare_phyla$total_spp_common_phyla_collected, big.mark=",")`) for the `r n_idigbio_phyla - records_rare_phyla$n_rare_phyla` other phyla. Yet, these undersampled taxa, especially those of small body size, are often the most abundant, and are likely to play an important role in the detection of changes in environmental quality[@Bik2012]. Additionally, the ecological implications of changes in the structure and composition of the communities of these small-bodied organisms, on higher trophic levels and ecosystems are largely unexplored.

The proportion of samples in museum collections that are identified to species decreases substantially toward the present, and shows a marked decline around the 1970's (Fig. \@ref(fig:identification-through-time)), demonstrating limiting taxonomic effort and expertise.

@@

We extracted `r format(nrow(gom_worms_species), big.mark=",")` and `r format(nrow(kozloff_worms_species), big.mark=",")` marine invertebrate species from checklists of the Gulf of Mexico [@Tunnell2009] and ofx the Pacific Northwest [@Kozloff1996] respectively. For the Gulf of Mexico, biodiversity databases included more species for the same geographic area than the one used by the authors of the checklist (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_gom_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(n_distinct(idigbio_gom_records$worms_valid_name), big.mark=",")` species, $n =$  `r format(nrow(idigbio_gom_records), big.mark=",")` records; OBIS only: `r format(n_distinct(obis_gom_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_gom_records), big.mark=",")` records). For the PNW, fewer species than included in the regional checklist were recovered from biodiversity databases (iDigBio and OBIS combined: `r format(dplyr::n_distinct(combined_pnw_records$worms_valid_name), big.mark=",")` species; iDigBio only: `r format(dplyr::n_distinct(idigbio_pnw_records$worms_valid_name), big.mark=",")`  species, $n=$ `r format(nrow(idigbio_pnw_records), big.mark=",")` records; OBIS only `r format(dplyr::n_distinct(obis_pnw_records$worms_valid_name), big.mark=",")` species, $n=$ `r format(nrow(obis_pnw_records), big.mark=",")` records). The combination of the species listed in these three sources leads to `r format(n_distinct(total_gom_records$worms_valid_name), big.mark=",")` species for the GOM and `r format(n_distinct(total_pnw_records$worms_valid_name), big.mark=",")` species for the PNW. Conversely, about XX% of the total species records for the Gulf of Mexico, and XX% for the Pacific Northwest are missing from the checklists and represented only in databases.

@@

Several factors may contribute to species being included in the regional lists but not having records in biodiversity databases. The first category of biases is linked to the likelihood of having specimens for these species in museum collections. Rare species may only have been collected a few times, and the corresponding records have not been digitized yet. Species that are difficult to preserve because they are large, very small, or fragile, are underrepresented in collections. The other category of biases is associated with data quality for these records. Some of these species may not have been digitized, or lack geographic coordinates associated, and therefore would be missed by our queries. While it is difficult to assess biases associated with the likelihood of preservation of a species, we investigated how missing geographic coordinates affect the discrepancies between the checklists and the databases.

<!-- include bias in digitization -->

Missing geographic coordinates reduces the overlap between the list of species from the regional list and the databases. Using iDigBio, we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are completely absent from iDigBio, while of the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that do appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them. Additional digitization and georeferencing of collections is needed to improve biodiversity knowledge at regional scales. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into biodiversity information systems. Similarly, only a small fraction of the marine invertebrates records from institutions located on the Pacific coast has been digitized, as indicated by the paucity of Pacific coast collections listed in Table \@ref(tab:records-table).

Two factors most likely contribute to explanations of why some species are in databases but are not included in the regional checklists: (1) the species have been documented for the first time since the publication of the checklist, (2) the species records were not accessible or known to the authors, or (3) the species records are erroneaous and were excluded by the compilers of the checklist.

Recently recorded species contribute modestly to this discrepancy for the Gulf of Mexico. Only `r sum(recently_collected$gom_in_dbs$min_year > 2001)` species (`r format_output( sum(recently_collected$gom_in_dbs$min_year > 2001)/n_distinct(total_gom_records$worms_valid_name) * 100)`% of the total species richness, `r format_output(sum(recently_collected$gom_in_dbs$min_year > 2001)/spp_not_in_lists$gom_not_in_list * 100)`% of the species not on the checklist) are only represented by collections made since the publication of the checklist. For the older Pacific Northwest checklist, this bias is larger with  `r sum(recently_collected$pnw_in_dbs$min_year > 1996)` species (`r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/n_distinct(total_pnw_records$worms_valid_name) * 100)`% of the total species richness,  `r format_output(sum(recently_collected$pnw_in_dbs$min_year > 1996)/spp_not_in_lists$pnw_not_in_list * 100)`% of the species not on the checklist) collected subsequent to publication.

The authors of these checklists had no or limited access to digital databases that have since been developed and this likely contributed to species being missed.

<!-- GP says he moved this up, but I don't see it: Many of these records were difficult to discover before digitization which, for the most part, has occurred after the compilation of these checklists. We estimate that using biodiversity databases increases the number of species included in the checklist by `r format_output(spp_not_in_lists$gom_not_in_list/n_distinct(total_gom_species$worms_valid_name)*100)`% for the Gulf of Mexico, and `r format_output(spp_not_in_lists$pnw_not_in_list/n_distinct(total_pnw_species$worms_valid_name)*100)`% for the Pacific Northwest. -->

@@

Since its introduction in the early 2000s, DNA barcoding has revitalized invertebrate taxonomy by facilitating species identification and delineation [@Meyer2005-plos]. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of specimens in environmental samples (eDNA) cost effectively [@Leray2015], making it possible to estimate diversity and identify taxa from species-rich samples at unprecedented scales. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies are poised to revolutionize large-scale assessment and monitoring of biodiversity [@Bohan2017]. However, the utility of the data generated hinges on reliably identified DNA barcode sequences that can serve for matching and identification of eDNA transcripts [@Leray2015]. Thus insights from eDNA technologies depend on the level of completeness and accuracy of comparative, identified, species-level DNA barcode coverage.

The marine invertebrates of the US are genetically characterized at a relatively high level: `r format_output(prop_spp_barcoded$p_barcoded * 100)`% (n = `r format(nrow(dplyr::distinct(combined_species, worms_valid_name)), big.mark=",")`) of the digitized fauna has DNA barcodes available from specimens taken somewhere in the world. Coverage varies by phylum from `r format_output(stat_barcoding$min_eez_value * 100)`% (`r stat_barcoding$min_eez_phylum`) to `r format_output(stat_barcoding$max_eez_value * 100)`% (`r stat_barcoding$max_eez_phylum`) (Fig. \@ref(fig:proportion-barcoded)). The GOM has comparable DNA barcoding coverage to the US average, but the PNW is better barcoded with >30% species-level coverage across each of the five most diverse phyla. This is driven in part by the active Canadian barcoding efforts in British Columbia.  In 2012, it was estimated that, at a global scale, less than 3% of the marine species had been sequenced [@Appeltans2012], and that this number was even lower in some areas (e.g., 2.6% for the Antarctic [@Grant2009]).
